---
title: "**School Sanitation: Lessons from the COVID-19 Pandemic**"
subtitle: "**Global Trends** and **Insights** from **2020**"
author: "Emma Doogue"
date: "2024-03-31"
format: 
  html: 
      theme: zephyr
      backgroundcolor: lightyellow
      toc: true


execute: 
  echo: false
  warning: false
  message: false
---
## Introduction

In **2020**, the world faced unprecedented challenges due to the **COVID-19 pandemic**, affecting various sectors, including **education. Basic sanitation in schools** is a fundamental right and it is a necessity for safe learning environments, as a result of the **pandemic sanitation** in **schools** came under **scrutiny**. This report delves into the state of basic sanitation in schools across different countries across the world over the years, with a focus on the period of **2020**.

```{r}
#| label: setup
#| include: false

#libraries
library(tidyverse)
library(plotly)

#data
unicef_indicator_1 <- read_csv("unicef_indicator_1.csv")
unicef_metadata <- read_csv("unicef_metadata.csv")

#transformations
data_join <- unicef_metadata %>%
  full_join(unicef_indicator_1, by = c("country", "year" = "time_period"))
```

## Global Overview

```{r}
#| fig-cap: "Our journey begins with a **global map visualisation**, showcasing the proportion of schools with **basic sanitation** from **2000** to **2020**. This comprehensive view highlights both **progress and setbacks** across different regions, setting the stage for a deeper investigation into how the pandemic might have influenced these trends. This map illustrates the **evolving landscape of basic sanitation in schools** worldwide, highlighting **progress** in some regions, persistent disparities, and the critical importance of **sustained efforts** leading up to and during the **COVID-19 pandemic**. *Note: There was no data provided for the countries shaded in grey*."

map_world <- map_data("world")

map_data_join <- full_join(unicef_indicator_1, map_world, by = c("country" = "region"))

ggplot(map_data_join) + 
  aes(x = long, y = lat, group = group, fill = obs_value) + 
  geom_polygon() + 
  scale_fill_gradient(low = "blue", high = "orange", name = "%PBSS") + 
  labs(x = "Longitude", y = "Latitude", title = "Proportion of Schools with Basic Sanitation") + 
  theme_void() +
  theme(plot.background = element_rect(fill = "lightyellow"), axis.line = element_blank()) 
```


## A Decade in Review


```{r}
#| fig-cap: "Focusing on a select group of countries (**Germany, India, Niger, Mexico, and Afghanistan**), we observe the trajectory of **basic sanitation in schools** over the last decade. While **Germany** boasts a consistent **100%** rate, showcasing its robust infrastructure and commitment to student welfare, **Afghanistan** remains constant at **38.2%**, reflecting persistent challenges. In contrast, **Mexico** shows a decline in basic sanitation, raising concerns about deteriorating conditions. **India** and **Niger**, however, offer glimmers of hope with their upward trends, suggesting improvements in their educational environments. *Note: there is gaps in the data therefore leading to gaps in this time series.*"
 p <- data_join %>%
   filter(country %in% c("Germany", "India", "Niger", "Mexico", "Afghanistan"), year >= 2010 & year <= 2021) %>%
   ggplot(aes(x = year, y = obs_value, group = country, colour = country)) +
   geom_line() +
   labs(x = "Year", y = "Value", title = "Time Series by Country") + 
   theme(legend.position = "none") + 
   theme(panel.background = element_rect(fill = "lightyellow"), plot.background = element_rect(fill = "lightyellow"))

  
 ggplotly(p)
```

## The Impact of GDP on Sanitation in 2020

```{r}
#| fig-cap: "We then narrow down to the pivotal year of **2020**, we explore the relationship between a country's economic status (measured by **GDP**) and the proportion of schools with **basic sanitation**. Our scatterplot for selected countries with an average **GDP of 0-50,000 USD** reveals a complex picture. It prompts us to question how economic health influences a country's ability to provide basic sanitation in schools, especially during a global health crisis.The scatterplot focuses on countries with a **GDP between 0 and 50,000 USD** to explore how varying economic statuses within a broad but defined range influence the provision of basic sanitation in schools, particularly in the context of the **diverse challenges** and **priorities** faced by these **nations** in **2020**."
vis_3 <- data_join %>%
  filter(year == 2020) %>%
  group_by(country) %>%
  summarise(
    Avg_GDP_per_capita = mean(`GDP per capita (constant 2015 US$)`, na.rm = TRUE),
    Obs_value_avg = mean(obs_value, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Obs_value_avg, y = Avg_GDP_per_capita, color = country)) +
  geom_point() +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 50000)) +
  labs(
    y = "Average GDP per capita (constant 2015 US$)",
    x = "Average Obs Value",
    title = "Average GDP per Capita by Country for 2020"
  ) + theme(panel.background = element_rect(fill = "lightyellow"), plot.background = element_rect(fill = "lightyellow"))

ggplotly(vis_3)
```

## Comparative Analysis in 2020

```{r}
#| fig-cap: "The final piece of our narrative is a **bar chart** comparing the average proportion of schools with basic sanitation in **2020** among our selected countries. This barchart is visually highlighting **differences** between **countries** and underscoring the varying levels of access to **essential hygiene facilities** within these **educational environments**."
vis_4 <- data_join %>%
  filter(year == 2020, country %in% c("Germany", "India", "Niger", "Mexico", "Afghanistan")) %>%
  group_by(country) %>%
  summarise(Avg_obs_value = mean(obs_value, na.rm = TRUE)) %>%
  ggplot(aes(x = country, y = Avg_obs_value, fill = country)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(y = "Average Obs Value", x = "Country", title = "Average Obs Value by Country for 2020") +
  theme(legend.position = "none") +
  theme(panel.background = element_rect(fill = "lightyellow"), plot.background = element_rect(fill = "lightyellow"))

ggplotly(vis_4)
```

## Conclusion

The **COVID-19 pandemic** has underscored the critical importance of **basic sanitation in schools**, directly impacting students' health, well-being, and access to education. Our visualisations tell a story of varied responses and outcomes across different countries, influenced by **economic conditions, policy decisions**, and **infrastructure capabilities**. As we move forward, it becomes evident that ensuring basic sanitation in schools is not just a matter of health but a foundation for **educational equity** and **resilience** against future crises.

